The Pair Distribution Function of the Inhomogeneous Hard-Sphere Fluid

### 2014-01-21 papers/pair-correlation/paper.tex
edit of paper (very quick) Tested on bennet. Took 0:16.12.

The standard approach in liquid state theory is to treat a liquid as a
hard-sphere reference fluid with attractive interactions that are
treated perturbatively~\cite{hansen2006theory}.  Recently there have
been considerable advances in the development of classical density
functionals based on the
application of these theoretical approaches to inhomogeneous
densities~\cite{ jain2007modified, gloor2007prediction, gross2009density,
  %felipe2001examination, gloor2002saft,
  %gloor2004accurate, clark2006developing,
  kahl2008modified,
  % yu2002fmt-dft-inhomogeneous-associating,
  %fu2005vapor-liquid-dft, bryk2006density,
  hughes2013classical,
  %segura1998comparison, felipe2001examination,
  %gloor2002saft, gloor2004accurate, fu2005vapor-liquid-dft,
  bryk2006density, clark2006developing,
  kahl2008modified, gross2009density,
  %yu2002fmt-dft-inhomogeneous-associating,
  sundararaman2012computationally, marshall2013density}.  The
perturbation theory treatment of intermolecular interactions relies on
the pair distribution function of the reference fluid:
$g^{(2)}(\rr_1,\rr_2)$.  Unlike the radial distribution function of a
homogeneous fluid, there does not currently exist a tractable form for
the pair distribution function of an inhomogeneous hard-sphere fluid,
suitable for use in constructing a density
functional~\cite{gloor2007prediction, jain2007modified}.

@@

The classic (and earliest) approach for computing the pair
distribution function is to use Percus' trick of treating one sphere
as an external field, and using the resultant equilibrium density to
find the pair distribution function~\cite{hansen2006theory}.  This
elegant approach lends itself to DFT, and can be used to compute and
plot the pair distribution function, but requires a full free-energy
minimization \emph{for each position} $\rr_1$ in
$g^{(2)}(\rr_1,\rr_2)$, and hence would be prohibitively expensive as
a tool in constructing a free energy functional.

The first canonical inhomogeneous configuration for the hard-sphere
fluid is the system consisting of a hard sphere at a hard wall.  In
1986, Plischke and Henderson solved the pair distribution function of
this system using integral equation theory under the Percus-Yevick
approximation~\cite{plischke1986pair}.  Ten years later,
G{\"o}tzelmann \emph{et al.} found the pair distribution function near
a hard wall using classical DFT by finding the direct correlation
function by taking a functional derivative of the free energy
functional and then solving the Ornstein-Zernicke
equation~\cite{gotzelmann1996structure}.  The approach of solving the
Ornstein-Zernicke equation for an inhomogeneous system is
computationally challenging~\cite{paul2003variational}.

A second inhomogeneous configuration that is of interest is the
test-particle configuration, in which one hard sphere is fixed in the
otherwise homogeneous fluid.  Where the hard-wall is a surface with no
curvature, the test-particle configuration has a surface with
curvature at the molecular length scale.  In this case, the density
gives the pair distribution function---this is just Percus'
trick---and the pair distribution function gives the triplet
distribution function.  The triplet distribution function was computed
by Gonz\'alez \emph{et al.} using the test-particle approach with
\emph{two} spheres fixed~\cite{gonzalez1999test}.

Lado recently introduced a new and improved efficient algorithm for
implementing integral equation theory for inhomogeneous fluids, which
computes $g^{(2)}(\rr_1,\rr_2)$~\cite{lado2009efficient}.  While this
approach is two orders of magnitude more efficient than previous
implementations, it remains a computationally expensive approach, and
unsuitable for repeated evaluation within a free-energy minimization.

@@

Approaches that do not diverge use an average of the density over some
volume. Fischer and Methfessel~\cite{fischer1980born} introduce the
approximation
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is one of the fundamental measures defined in Fundamental
Measure Theory~\cite{rosenfeld1989free}, which is an integral of the
density over a spherical volume:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
where $\sigma$ is the hard-sphere diameter.  Equation~\ref{eq:fischer}
is computationally awkward, because it relies on the midpoint
$\tfrac12(\rr_1+\rr_2)$.  Moreover, the approach of Fischer and
Methfessel is intended to approximate the pair distribution function only
at contact, when the distance between $\rr_1$ and $\rr_2$ is a hard-sphere
diameter.

@@

@@

In the previous approaches mentioned above, function forms of radial
distributions (used to approximate pair correlation) were dependent
upon independent variables $\eta$, the density, whether at single
points or averaged over sourrounding volumes.  We can achieve more
accuracy by using instead a function dependent upon our $g_{\sigma}$,
which holds more information about an inhomogeneous system
then an average $\eta$ does.  We therefore need a function form for
our radial distribution that is dependent upon $g_{\sigma}$ and that
will approach the correct solution in the homogeneous limit and also
in the limit of contact distance between the two points.

@@

Figure~\ref{fig:pair-distribution-1}b
and~\ref{fig:pair-distribution-3}b show the pair distribution function
as plotted along a path illustrated in
Fig.~\ref{fig:pair-distribution-1}a.  These plots compare this work
with Monte Carlo results, as well as the approximations of Sokolowski
and Fischer~\cite{sokolowski1992role} and that of Fischer and
Methfessel~\cite{fischer1980born} at packing fractions of 0.1 and 0.3.
The approach of Fischer and Methfessel is only defined when the two
spheres are in contact, and is therefore only plotted on that segment
of the path.  As an input to both previous approximations we use the
hard sphere radial distribution function found with Monte Carlo
simulation, interpolated as necessary.  We find that both previous
approximations to the pair distribution function predict the
qualtitative angular dependence of the pair distribution function at
contact better than this work.  However, in each case the pair
distribution function has a systematic error at contact---either too
high or too low.  Thus while our approximation is smoother than either
of the existing approaches, its errors will have a tendancy to cancel
when used in a perturbation expansion.  At higher densities, the
approximation of Fischer and Methfessel requires evaluating the radial
distribution function at densities significantly higher than the
freezing density, which poses numerical difficulties when using the
radial distribution function from simulation.  When the two points are
more than a radius away from contact, we find that any of these
approaches gives a reasonable prediction, with our approach
underestimating the oscillations in $g^{(2)}$, as expected based on our
fit in Fig.~\ref{fig:radial-distribution}.

@@

Figures~\ref{fig:triplet-contact-distribution-3}-\ref{fig:triplet-inbetween-distribution-4}
compare the triplet distribution function computed using our
approximation to $g^{(2)}$ with results from Monte Carlo simulation.
In each figure, $\rr_1$ and $\rr_2$ are held fixed, while the third
position $\rr_3$ is varied.  The test-particle position in each case
is $\rr_1$, which is on the left-hand side of the figure.  In the
color plot on the left, the first and second spheres are displayed in
grey.  On the bottom half of each left-hand plot is the triplet
correlation function as computed using the approximation presented in
this work.  The top half displays the triplet correlation function
from Monte Carlo.  On the right, we display \fixme{this needs to be finished...}

>>>>

The standard approach in liquid state theory is to treat a liquid as a
hard-sphere reference fluid with attractive interactions that are
treated perturbatively~\cite{hansen2006theory}.  Recently there have
been considerable advances in the development of classical density
functionals based on the
application of these theoretical approaches to inhomogeneous
densities~\cite{ jain2007modified, gloor2007prediction, gross2009density,
  %felipe2001examination, gloor2002saft,
  %gloor2004accurate, clark2006developing,
  kahl2008modified,
  % yu2002fmt-dft-inhomogeneous-associating,
  %fu2005vapor-liquid-dft, bryk2006density,
  hughes2013classical,
  %segura1998comparison, felipe2001examination,
  %gloor2002saft, gloor2004accurate, fu2005vapor-liquid-dft,
  bryk2006density, clark2006developing,
  kahl2008modified, gross2009density,
  %yu2002fmt-dft-inhomogeneous-associating,
  sundararaman2012computationally, marshall2013density}.  The
perturbation theory treatment of intermolecular interactions relies on
the pair distribution function of the reference fluid:
$g^{(2)}(\rr_1,\rr_2)$.  Unlike the radial distribution function of a
homogeneous fluid, there does not currently exist a tractable form for
the pair distribution function of an inhomogeneous hard-sphere fluid,
suitable for use in constructing a density
functional~\cite{gloor2007prediction, jain2007modified}. \fixme{We need to say what a density functional is}

@@

The classic (and earliest) approach for computing the pair
distribution function is to use Percus' trick of treating one sphere
as an external field, and using the resultant equilibrium density to
find the pair distribution function~\cite{hansen2006theory}.  This
elegant approach lends itself to DFT \fixme{Define DFT earlier!}, and can be used to compute and
plot the pair distribution function, but requires a full free-energy
minimization \emph{for each position} $\rr_1$ in
$g^{(2)}(\rr_1,\rr_2)$, and hence would be prohibitively expensive as
a tool in constructing a free energy functional.

The canonical inhomogeneous configuration for the hard-sphere
fluid is the system consisting of a hard sphere at a hard wall.  In
1986, Plischke and Henderson solved the pair distribution function of
this system using integral equation theory under the Percus-Yevick
approximation~\cite{plischke1986pair}.  Ten years later,
G{\"o}tzelmann \emph{et al.} found the pair distribution function near
a hard wall using classical DFT by finding the direct correlation
function by taking a functional derivative of the free energy
functional and then solving the Ornstein-Zernicke
equation~\cite{gotzelmann1996structure}.  The approach of solving the
Ornstein-Zernicke equation for an inhomogeneous system is
computationally challenging~\cite{paul2003variational}.

Another inhomogeneous configuration that is of interest is the
test-particle configuration, in which one hard sphere is fixed in the
otherwise homogeneous fluid.  Where the hard-wall is a surface with no
curvature, the test-particle configuration has a surface with
curvature at the molecular length scale.  In this case, the density
gives the pair distribution function---this is just Percus'
trick---and the pair distribution function gives the triplet
distribution function.  The triplet distribution function has been computed
by Gonz\'alez \emph{et al.} using the test-particle approach with
\emph{two} spheres fixed~\cite{gonzalez1999test}.

Lado recently introduced a new and improved efficient algorithm for
implementing integral equation theory for inhomogeneous fluids, which
computes $g^{(2)}(\rr_1,\rr_2)$~\cite{lado2009efficient}.  While this
approach is two orders of magnitude more efficient than previous
implementations, it remains a computationally expensive approach, and
unsuitable for repeated evaluation within a free-energy minimization.
\fixme{When we introduce DFT, we should talk about minimization.}

@@

Approaches that do not diverge use an average of the density over some
volume. Fischer and Methfessel~\cite{fischer1980born} introduce the
approximation
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is one of the fundamental measures defined in Fundamental
Measure Theory~\cite{rosenfeld1989free}, which is an integral of the
density over a spherical volume:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
where $\sigma$ is the hard-sphere diameter.  Equation~\ref{eq:fischer}
is computationally awkward, because it relies on the midpoint
$\tfrac12(\rr_1+\rr_2)$.  Moreover, the approach of Fischer and
Methfessel is intended to approximate the pair distribution function only
at contact, when the distance between $\rr_1$ and $\rr_2$ is the hard-sphere
diameter.

@@

\fixme{Add a paragraph here explaining the work of Schulte {\em et
    al.}  In particular, we want to emphasize that we have an accurate
representation for an \emph{average} of the pair distribution function
\emph{at contact}.  Then we can cite this in the ``Our Approach''
section.}

@@

In the approaches mentioned above, function forms of radial
distributions (used to approximate pair correlation) were dependent
upon independent variables $\eta$, the density, whether at single
points or averaged over sourrounding volumes.  We can achieve greater
accuracy by using instead a function dependent upon our $g_{\sigma}$,
which holds more information about an inhomogeneous system
then a simple convolution of the density does.  We therefore seek a function form for
our radial distribution that is dependent upon $g_{\sigma}$ and that
will approach the correct solution in the homogeneous limit and also
in the limit of contact distance between the two points.

@@

Figure~\ref{fig:pair-distribution-1}b
and~\ref{fig:pair-distribution-3}b show the pair distribution function
as plotted along a path illustrated in
Fig.~\ref{fig:pair-distribution-1}a.  These plots compare this work
with Monte Carlo results, as well as the approximations of Sokolowski
and Fischer~\cite{sokolowski1992role} and that of Fischer and
Methfessel~\cite{fischer1980born} at packing fractions of 0.1 and 0.3.
The approach of Fischer and Methfessel is only defined when the two
spheres are in contact, and is therefore only plotted on that segment
of the path.  As an input to both previous approximations we use the
hard sphere radial distribution function found with Monte Carlo
simulation, interpolated as necessary.  We find that both previous
approximations to the pair distribution function predict the
qualtitative angular dependence of the pair distribution function at
contact better than this work.  However, in each case the pair
distribution function has a systematic error at contact---either too
high or too low.  Thus while our approximation is smoother than either
of the existing approaches, its errors will have a tendancy to cancel
when used in a perturbation expansion.  At higher densities, the
approximation of Fischer and Methfessel requires evaluating the radial
distribution function at densities significantly higher than the
freezing density, which poses numerical difficulties when using the
radial distribution function from simulation.  When the two points are
more than a radius away from contact, we find that any of these
approaches gives a reasonable prediction, with our approach
underestimating the oscillations in $g^{(2)}$, as expected based on the
fit in Fig.~\ref{fig:radial-distribution}.

@@

\fixme{The following paragraph needs to be rewritten when we've
  reorganized the figures. (Obviously.)}

Figures~\ref{fig:triplet-contact-distribution-3}-\ref{fig:triplet-inbetween-distribution-4}
compare the triplet distribution function computed using our
approximation to $g^{(2)}$ with results from Monte Carlo simulation.
In each figure, $\rr_1$ and $\rr_2$ are held fixed, while the third
position $\rr_3$ is varied.  The test-particle position in each case
is $\rr_1$, which is on the left-hand side of the figure.  In the
color plot on the left, the first and second spheres are displayed in
grey.  On the bottom half of figures \fixme{b,d,etc?} is the triplet
correlation function as computed using the approximation presented in
this work.  The top half of those figures displays the triplet
correlation function from Monte Carlo.  Figures~\fixme{a,c,e?} show
the triplet distribution function along the paths illustrated in the
corresponding 2D plots.

Figures~\fixme{which} and~\fixme{which} show the triplet distribution
function when two of the positions $\rr_1$ and $\rr_2$ are in contact.
The triplet distribution function predicted by this work is slightly
asymmetric, indicating a failure in thermodynamic consistency.  Once
again, we observe the lack of oscillations at large distances.
Figures~\fixme{which} and~\fixme{which} show the triplet distribution
function when two of the positions $\rr_1$ and $\rr_2$ are spaced so
that the third could just fit between them.  Again, the triplet
distribution function predicted by this work is slightly asymmetric,
and lacks the oscillations seen in the Monte Carlo simulations at
large distances.  In both cases, our agreement with simulation is
comparable to that of previous approximations.
